---
title: "ML ASS 1 LR CVD PREDICTION NEW"
author: "KIBIDI"
date: "2026-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Course: PROFFESIONAL CERTIFICATE IN DATA ANALYSIS,  MACHINE LEARNING AND AI
Instructor: Lumumba W. Victor
Email: lumumbavictor172@gmail.com

student: KIBIDI LERENCER
Reg no: CDAM/0047/2026
Email: kabeyi24leren@gmail.com

Date Issued: April , 2026

```{r}
 setwd("C:/Users/h6p/Desktop/ASSIGNMENTS")
```

3.1 Task 1: Data Import and Exploration (10 Marks)

#3.1.1. Load all necessary libraries 
```{r}
library(tidyverse)# Data manipulation
library(caret)# Unified ML interface
library(pROC)# ROC analysis
library(e1071)# SVM
library(class)# KNN
library(randomForest) # Random Forest
library(dplyr)
```


#3.1.2. Import the dataset medical_data.csv.
```{r}
pressure <- read_csv("medical_data.csv")
head(pressure)
```



#3.1.3a. Display the structure of the data 
```{r}
str(pressure)
```



#3.1.3b.summary statistics 
```{r}
summary(pressure)
```

#3.1.4a. Check for missing values and report the count per variable.
```{r}
colSums(is.na(pressure))
```


#3.1.4b) report the count and percentage per variable.
```{r}
missing_summary <- data.frame(
  Variable = names(pressure),
  Missing_Count = colSums(is.na(pressure)),
  Missing_Percent = colMeans(is.na(pressure)) * 100
)

missing_summary
```


#3.1.4c)report sum of duplicates
```{r}
sum(duplicated(pressure))
```


#3.1.5a. Visualize the distribution of the target variable CVD using a bar plot.
```{r}
ggplot(pressure, aes(x = CVD, fill = CVD)) +
  geom_bar() +
  geom_text(stat = "count", 
            aes(label = after_stat(count)), 
            vjust = -0.5, size = 5) +
  labs(title = "Bar Plot of Cardiovascular Disease", 
       x = "Cardiovascular Disease Status (0 = No, 1 = Yes)", 
       y = "Count") +
  theme_bw() +
  expand_limits(y = max(table(pressure$CVD)) * 1.2)  # ← Extend y-axis by 20%
```



#3.1.5b,visualize ouliers via box plots
```{r}
boxplot(pressure$age, main="Outliers_age")
boxplot(pressure$gender, main="Outliers_gender")
boxplot(pressure$bmi, main="Outliers_bmi")
boxplot(pressure$blood_pressure, main="Outliers_blood_pressure")
boxplot(pressure$cholesterol, main="Outliers_cholesterol")
boxplot(pressure$glucose, main="Outliers_glucose")
boxplot(pressure$smoking, main="Outliers_smoking")
boxplot(pressure$alcohol, main="Outliers_alcohol")
boxplot(pressure$physical_activity, main="Outliers_physical_activity")
boxplot(pressure$family_history, main="Outliers_family_history")
boxplot(pressure$stress_level, main="Outliers_stress_level")
boxplot(pressure$heart_rate, main="Outliers_heart_rate")
boxplot(pressure$sleep_hours, main="Outliers_sleep_hours")
boxplot(pressure$medication_adherence, main="Outliers_medication_adherence")
boxplot(pressure$urban_residence, main="Outliers_urban_residence")
```




3.2 Task 2: Data Preprocessing (15 Marks)
#3.2.1. Handle missing values (justify your method: removal orimputation).
```{r}
pressure_clean <- na.omit(pressure)
```
Justification: Complete case analysis  to ensure modeltraining uses only verified patient records, training uses only verified patient records. If NA rate >5%, multiple imputation would be preferred in production.



#3.2.2. Convert categorical variables (Gender, Smoking, Alcohol, Exercise, FamilyHistory,Medication) to factors.
```{r}
categorical_vars <- c("gender", "smoking", "alcohol","physical_activity",
                      "family_history", "medication_adherence")

pressure_clean[categorical_vars] <- lapply(pressure_clean[categorical_vars], as.factor)
```


#3.2.3. Ensure the target variable CVD is a factor with levels "No" (0) and "Yes" (1).
```{r}
pressure_clean$CVD <- factor(pressure_clean$CVD,
                                 levels = c("No", "Yes"))
```


#3.2.4. Scale numeric predictors (mean = 0, sd = 1) using preProcess() from caret.
#3.2.4a) site numerical variables
```{r}
numeric_vars <- names(pressure_clean)[sapply(pressure_clean, is.numeric)]
numeric_vars
```



3.3 Task 3: Data Partitioning (10 Marks)

#3.3.1. Set the seed to 2026 for reproducibility.
#3.3.2. Split the data into 70% training and 30% testing using stratified sampling (to preserve class balance).Note: Fit the scaling parameters on the training data only
```{r}
# Set seed for reproducibility
set.seed(2026)

#Stratified 70/30 split
train_index <- createDataPartition(pressure_clean$CVD, p = 0.7, list = FALSE)
train_data  <- pressure_clean[train_index, ]
test_data   <- pressure_clean[-train_index, ]

#Scale numeric predictors (fit parameters on TRAINING data ONLY)
preproc_params <- preProcess(train_data[, numeric_vars], method = c("center", "scale"))
train_data[, numeric_vars] <- predict(preproc_params, train_data[, numeric_vars])
test_data[, numeric_vars]  <- predict(preproc_params, test_data[, numeric_vars])
```


#3.2.4c)verify scaling using numerical variable 'age'
```{r}
cat("\n========== SCALING VERIFICATION ==========\n")
cat("Original age - Mean:", mean(pressure_clean$age), "| SD:", sd(pressure_clean$age), "\n")
cat("Training age (scaled) - Mean:", round(mean(train_data$age), 3), 
    "| SD:", round(sd(train_data$age), 3), "\n")
cat("Test age (scaled) - Mean:", round(mean(test_data$age), 3), 
    "| SD:", round(sd(test_data$age), 3), "\n")

cat("\nAll scaled variables (training data):\n")
print(round(colMeans(train_data[, numeric_vars]), 3))
print(round(apply(train_data[, numeric_vars], 2, sd), 3))
```

#3.3.3. Report the class distribution in both training and testing sets.
#Report the class distribution in both training and testing sets
```{r}
dim(train_data)
dim(test_data)

cat("\n--- Training Set Distribution ---\n")
print(prop.table(table(train_data$CVD)))
cat("\n--- Test Set Distribution ---\n")
print(prop.table(table(test_data$CVD)))
```

#3.2.4d. verify split sampling
```{r}
prop.table(table(train_data$CVD))
prop.table(table(test_data$CVD))
```



#3.2.4e) Check stratification worked
```{r}
# Check stratification worked
cat("\nStratification check - CVD proportion:\n")
cat("Original:", round(prop.table(table(pressure_clean$CVD))[2] * 100, 1), "%\n")
cat("Training:", round(prop.table(table(train_data$CVD))[2] * 100, 1), "%\n")
cat("Testing:", round(prop.table(table(test_data$CVD))[2] * 100, 1), "%\n")
```

3.4 Task 4: Model Training with trainControl (20 Marks)
#3.4.1. Define a trainControl scheme with 10-fold cross-validation.
```{r}
ctrl <- trainControl(
  method = "cv", 
  number = 10, 
  classProbs = TRUE, 
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)
```


#3.4.2a. Train a Logistic Regression model using caret::train() with method "glm".
```{r}
set.seed(2026)

model_lr <- train(
  CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + family_history + stress_level + sleep_hours + urban_residence,
  

  data = train_data,                   # Logistic regression
  method = "glm",                     # Binary classification             
  trControl = ctrl,
  family = binomial
)

print(model_lr)
```

#3.4.2b) summary results
```{r}
model_lr$results
```


#3.4.4a. Extract & display coefficients
```{r}
# Extract coefficients
coef_vec <- coef(model_lr$finalModel)

# Convert to data frame
coef_df <- data.frame(
  Variable = names(coef_vec),
  Coefficient = as.numeric(coef_vec)
)

# Order from largest to smallest
coef_df <- coef_df[order(coef_df$Coefficient, decreasing = TRUE), ]

# Display
cat("\n--- Coefficients (Largest to Smallest) ---\n")
print(coef_df)
```
```{r}
#Extract & prepare
coef_vec <- coef(model_lr$finalModel)
coef_df <- data.frame(
  Variable = names(coef_vec),
  Coefficient = as.numeric(coef_vec)
)

# Remove intercept
coef_df <- subset(coef_df, Variable != "(Intercept)")

# Order by importance
coef_df <- coef_df[order(abs(coef_df$Coefficient), decreasing = TRUE), ]

# Plot
ggplot(coef_df, aes(x = reorder(Variable, Coefficient), y = Coefficient)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Logistic Regression Coefficients",
       x = "Variables",
       y = "Coefficient Value")
```



#3.4.3. Alternatively, fit a baseline model using base R glm() function.
```{r}
baseline_model <- glm(
  CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + family_history + stress_level + sleep_hours + urban_residence,
                      data = train_data, # Training data
                      family = binomial) # Logistic regression

# Display model summary
summary(baseline_model)
```

#3.4.4b. Extract and report the model coefficients.
```{r}
cat("\n--- Model Coefficients ---\n")
print(coef(baseline_model))
```





3.5 Task 5: Model Evaluation (25 Marks)
#3.5.1. Predict the class and probabilities on the test set.
```{r}
pred_class <- predict(model_lr, newdata = test_data)
pred_probs <- predict(model_lr, newdata = test_data, type = "prob")

pred_probs
```



#3.5.2. Generate a Confusion Matrix and report:
•Accuracy•Sensitivity (Recall)•Specificity•Kappa
```{r}
cm <- confusionMatrix(pred_class, test_data$CVD, positive = "Yes")
cat("\n--- Confusion Matrix Metrics ---\n")
cat(sprintf("Accuracy:   %.4f\n", cm$overall["Accuracy"]))
cat(sprintf("Sensitivity (Recall): %.4f\n", cm$byClass["Sensitivity"]))
cat(sprintf("Specificity: %.4f\n", cm$byClass["Specificity"]))
cat(sprintf("Kappa:      %.4f\n", cm$overall["Kappa"]))
print(cm)
```

#3.5.3. Plot the ROC Curve and calculate the AUC (Area Under Curve).
```{r}
roc_obj <- roc(test_data$CVD, pred_probs$Yes, levels = c("No", "Yes"), direction = "<")
auc_val <- auc(roc_obj)
cat("\nAUC (Area Under Curve):", round(auc_val, 4), "\n")

plot(roc_obj, main = "ROC Curve for CVD Logistic Regression", 
     col = "steelblue", lwd = 2, print.auc = TRUE, auc.polygon = TRUE, 
     auc.polygon.col = "skyblue")
abline(a = 0, b = 1, lty = 2, col = "gray")

```

#3.5.4a. Interpret the AUC value (e.g., poor, fair, good, excellent).



#3.5.4b Threshold Optimization
```{r}
thresholds <- seq(0.10, 0.50, by = 0.05)
thresh_results <- data.frame()

for(t in thresholds) {
  pred_t <- factor(ifelse(pred_probs$Yes > t, "Yes", "No"), levels = c("No", "Yes"))
  cm_t   <- confusionMatrix(pred_t, test_data$CVD, positive = "Yes")
  thresh_results <- rbind(thresh_results, data.frame(
    Threshold   = t,
    Sensitivity = cm_t$byClass["Sensitivity"],
    Specificity = cm_t$byClass["Specificity"],
    Accuracy    = cm_t$overall["Accuracy"],
    Kappa       = cm_t$overall["Kappa"]
  ))
}

print(thresh_results)
```


#3.5.4c. Visualize threshold trade-off
```{r}
thresh_results %>%
  pivot_longer(cols = c(Sensitivity, Specificity), names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = Threshold, y = Value, color = Metric)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Sensitivity vs. Specificity Across Classification Thresholds",
       x = "Probability Threshold", y = "Metric Value") +
  theme_minimal()
```


3.6 Task 6: Interpretation (20 Marks)

#3.6.2a. Calculate the Odds Ratios (OR) for these predictors (OR = eβ).
```{r}
# Extract Coefficients, P-values & Odds Ratios
coefs_df <- summary(model_lr$finalModel)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(Odds_Ratio = exp(Estimate)) %>%
  arrange(`Pr(>|z|)`)

print(coefs_df)
```

#3.6.2b. visualize Odds Ratios (OR) for these predictors (OR = eβ).
```{r}
odds_df <- summary(model_lr$finalModel)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(Odds_Ratio = exp(Estimate)) %>%
  select(Predictor, Odds_Ratio) %>%
  arrange(desc(Odds_Ratio))   # largest → smallest


ggplot(odds_df, aes(x = reorder(Predictor, Odds_Ratio), y = Odds_Ratio)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Odds Ratios (Ordered)",
       x = "Predictors",
       y = "Odds Ratio")
```



#3.6.1a. Select three significant predictors from your model.
```{r}
top_3 <- coefs_df %>% slice(1:3)
cat("\n--- Top 3 Significant Predictors ---\n")
print(top_3)
```


#3.6.1b. visualize top predictord of CVD
```{r}
# Extract & prepare odds ratios from your model
top_preds <- summary(model_lr$finalModel)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "(Intercept)") %>%
  mutate(OR = exp(Estimate)) %>%
  slice_min(order_by = `Pr(>|z|)`, n = 5)  # Top 5 by p-value

# Simple horizontal bar plot
ggplot(top_preds, aes(x = reorder(Predictor, OR), y = OR)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Top 5 CVD Predictors (Odds Ratios)",
       x = "", y = "Odds Ratio") +
  theme_minimal()
```


#3.6.3. Interpret the Odds Ratios in the context of CVD risk (e.g., "For every 1-unit increase in BMI, the odds of CVD increase by X%").
"• For every 1-hour increase in sleep, the odds of CVD decrease by 54.5% (OR=0.455), holding other factors constant.
• For every 1-unit increase in glucose, the odds of CVD increase by 106.7% (OR=2.07).
• For every 1-unit increase in stress level, the odds of CVD increase by 82.2% (OR=1.82)."


#3.6.4. Discuss any limitations of your model(assumptions,potentialbiases)
#3.6.4a. Class Imbalance 
If CVD positive cases are <20%, accuracy can be misleading. Kappa and Sensitivity are more reliable here.


#3.6.4b Linearity Assumption
Logistic regression assumes a linear relationship between log-odds and predictors. Non-linear effects (e.g., age vs. CVD risk) may require polynomial terms or splines.

#3.6.4c Correlation/Multicollinearity
Variables like SBP, DBP, and Cholesterol may be correlated. VIF checks or PCA could improve robustness.

#3.6.4d. Generalizability
Model performance depends on the dataset's demographic and clinical representativeness. External validation is recommended before clinical deployment.


#3.7.1 Correlation Heatmap (Numeric Predictors)
```{r}
library(corrplot)
numeric_df <- pressure_clean %>% select(where(is.numeric))
cor_mat <- cor(numeric_df, use = "complete.obs")

corrplot(cor_mat, method = "color", type = "upper", 
         tl.cex = 0.8, tl.srt = 45, addCoef.col = "black",
         title = "Correlation Matrix of Numeric Predictors", mar = c(0,0,2,0))
```



#3.7.3 Logistic Regression Coefficients (95% CI)
```{r}
# Extract coefficients + 95% CI
coef_plot <- data.frame(
  Predictor = names(coef(model_lr$finalModel))[-1],  # Remove intercept
  Estimate = coef(model_lr$finalModel)[-1],
  SE = summary(model_lr$finalModel)$coefficients[-1, 2]
) %>%
  mutate(
    Lower = Estimate - 1.96 * SE,
    Upper = Estimate + 1.96 * SE
  ) %>%
  arrange(Estimate)

ggplot(coef_plot, aes(x = reorder(Predictor, Estimate), y = Estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = Lower, ymax = Upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  coord_flip() +
  labs(title = "Logistic Regression Coefficients (95% CI)",
       x = "", y = "Log-Odds Coefficient") +
  theme_minimal()
```


#3.7.4 Predictor Importance Ranking (Absolute Coefficients)
```{r}
coef_importance <- data.frame(
  Predictor = names(coef(model_lr$finalModel))[-1],
  Abs_Coef = abs(coef(model_lr$finalModel)[-1])
) %>%
  arrange(desc(Abs_Coef)) %>%
  slice(1:8)  # Top 8

ggplot(coef_importance, aes(x = reorder(Predictor, Abs_Coef), y = Abs_Coef)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Predictors by Absolute Coefficient Size",
       x = "", y = "|Coefficient|") +
  theme_minimal()
```

