---
title: "ML ASS 2 CVD PREDICTION MODEL2"
author: "KIBIDI"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Course: PROFFESIONAL CERTIFICATE IN DATA ANALYSIS,  MACHINE LEARNING AND AI
Instructor: Lumumba W. Victor
Email: lumumbavictor172@gmail.com

student: KIBIDI LERENCER
Reg no: CDAM/0047/2026
Email: kabeyi24leren@gmail.com

Date Issued: April , 2026

```{r}
setwd("C:/Users/h6p/Desktop/ASSIGNMENTS")
```



3 Tasks
3.1 Task 1: Data Import and Exploration (10 Marks)

#3.1.1. Load all necessary libraries 
```{r}
library(tidyverse)# Data manipulation
library(caret)# Unified ML interface
library(pROC)# ROC analysis
library(e1071)# SVM
library(class)# KNN
library(randomForest) # Random Forest
library(dplyr)
library(vip)
```



#3.1.2. Import the dataset medical_data.csv.
```{r}
hypertension <- read_csv("medical_data.csv")
head(hypertension)
```

#3.1.3a) Display the structure of the data (str()) 
```{r}
(str(hypertension)) 
```



#3.1.3b) Display summary statistics (summary()).
```{r}
(summary(hypertension))
```




#3.1.4a) Check for missing values  per variable.
```{r}
colSums(is.na(hypertension))
```


#3.1.4b) report the count and percentage per variable.
```{r}
missing_summary <- data.frame(
  Variable = names(hypertension),
  Missing_Count = colSums(is.na(hypertension)),
  Missing_Percent = colMeans(is.na(hypertension)) * 100
)

missing_summary
```


#3.1.5. Visualize the distribution of the target variable CVD using a bar plot.
```{r}
 ggplot(hypertension, aes(x = factor(CVD))) +
geom_bar(fill = "steelblue") +
labs(title = "CVD Distribution", x = "CVD", y = "Count")
```



#3.1.6a) Calculate and report the class imbalance ratio
```{r}
class_counts <- table(hypertension$CVD)

IR <- max(class_counts) / min(class_counts)

IR

#percentage
prop.table(class_counts) * 100
```


#3.1.6b)visualize class imbalance ratio in percentage
```{r}
hypertension %>%
  count(CVD) %>%
  mutate(Percent = n / sum(n) * 100) %>%
  ggplot(aes(x = CVD, y = n, fill = CVD)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percent,1), "%")),
            vjust = -0.5) +
  labs(title = "CVD Class Distribution with Percentages",
       x = "CVD",
       y = "Count") +
  theme_minimal()
```



#3.1.3c)Frequency of variables m
```{r}
table(hypertension$gender)
table(hypertension$smoking)
table(hypertension$alcohol)
table(hypertension$family_history)
table(hypertension$medication_adherence)
table(hypertension$urban_residence)
table(hypertension$CVD)
```





3.2 Task 2: Data Preprocessing (15 Marks)
#3.2.1. Handle missing values (justify your method: removal or imputation).
```{r}
hypertension_clean <- na.omit(hypertension)
```

Missing values were handled using removal because the dataset contained no significant missing data,this is to  ensure data completeness without introducing bias from imputation.



#3.2.2. Convert categorical variables to factors:
```{r}
categorical_vars <- c("gender", "smoking", "alcohol",
                      "family_history", "medication_adherence", "urban_residence")

hypertension_clean[categorical_vars] <- lapply(hypertension_clean[categorical_vars], as.factor)
```


#3.2.3a) Convert the target variable CVD to a factor with levels "No" (0) and "Yes" (1).
```{r}
hypertension_clean$CVD <- factor(hypertension_clean$CVD,
                                 levels = c("No", "Yes"))
```

#3.2.5b.outlier visualization
```{r}
boxplot(hypertension_clean$age, main="Outliers_age")
boxplot(hypertension_clean$gender, main="Outliers_gender")
boxplot(hypertension_clean$bmi, main="Outliers_bmi")
boxplot(hypertension_clean$blood_pressure, main="Outliers_blood_pressure")
boxplot(hypertension_clean$cholesterol, main="Outliers_cholesterol")
boxplot(hypertension_clean$glucose, main="Outliers_glucose")
boxplot(hypertension_clean$smoking, main="Outliers_smoking")
boxplot(hypertension_clean$alcohol, main="Outliers_alcohol")
boxplot(hypertension_clean$physical_activity, main="Outliers_physical_activity")
boxplot(hypertension_clean$family_history, main="Outliers_family_history")
boxplot(hypertension_clean$stress_level, main="Outliers_stress_level")
boxplot(hypertension_clean$heart_rate, main="Outliers_heart_rate")
boxplot(hypertension_clean$sleep_hours, main="Outliers_sleep_hours")
boxplot(hypertension_clean$medication_adherence, main="Outliers_medication_adherence")
boxplot(hypertension_clean$urban_residence, main="Outliers_urban_residence")
```


# Function to remove outliers using IQR from a single column
```{r}
remove_outliers_iqr <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(x >= lower_bound & x <= upper_bound)  # Returns TRUE for non-outliers
}

# Apply to all numeric columns in your dataset
numeric_cols <- names(hypertension_clean)[sapply(hypertension_clean, is.numeric)]

# Create logical vector for rows to KEEP (no outliers in ANY column)
keep_rows <- rep(TRUE, nrow(hypertension_clean))

for(col in numeric_cols) {
  keep_rows <- keep_rows & remove_outliers_iqr(hypertension_clean[[col]])
}

# Remove rows with outliers
hypertension_cleant <- hypertension_clean[keep_rows, ]

cat("Original rows:", nrow(hypertension_clean))
cat("Rows after outlier removal:", nrow(hypertension_cleant))
cat("Rows removed:", nrow(hypertension_clean) - nrow(hypertension_cleant))
```

N/B : hypertension_cleant DOES NOT HAVE OUTLIERS

#3.2.3b)verify factors conversion
```{r}
str(hypertension_cleant)
```

#3.2.4. For KNN and SVM:Scale numeric predictors(mean=0, sd=1)using preProcess()from caret.
#3.2.4a) site numerical variables
```{r}
numeric_vars <- names(hypertension_cleant)[sapply(hypertension_cleant, is.numeric)]

numeric_vars
```
#3.2.4b)Scale numeric predictors
```{r}
 numeric_vars <- c("age", "bmi", "blood_pressure", "cholesterol",
                  "glucose", "stress_level", "heart_rate", "sleep_hours")

preproc <- preProcess(hypertension_cleant[, numeric_vars],
                      method = c("center", "scale"))

hypertension_scaled <- hypertension_cleant

hypertension_scaled[, numeric_vars] <- predict(preproc,
                                               hypertension_cleant[, numeric_vars])
```



#3.2.4c)verify scaling using numerical variable 'age'
```{r}
cat("\n========== SCALING VERIFICATION ==========\n")
cat("Original age - Mean:", mean(hypertension_cleant$age), "| SD:", sd(hypertension_clean$age), "\n")
cat("Scaled age - Mean:", round(mean(hypertension_scaled$age), 2), "| SD:", round(sd(hypertension_scaled$age), 2), "\n")
```

#3.2.5a. For Random Forest: Scaling is not required (tree-based).
Scaling is not required for Random Forest because it is a tree-based algorithm that splits data using thresholds rather than distance or magnitude. Therefore, changes in scale do not affect model performance or splits.



#3.2.6. Document your preprocessing decisions in the report.
handling missing values'
scalling
conversion to factors
removal of outliers



3.3 Task 3: Data Partitioning (10 Marks)

#3.3.1. Set the seed to 2026 for reproducibility.
```{r}
 set.seed(2026)
```


for the purpose of fare comparison i will partition scaled data and my train data will have no outliers and will also be scaled and i will use this acreoss all the 4 models


#3.3.2. Split the data into 70% training and 30% testing using stratified sampling (to preserve class balance).
```{r}
 train_index <- createDataPartition(hypertension_scaled$CVD, p = 0.70,
list = FALSE)
train_data <- hypertension_cleant[train_index, ]
test_data <- hypertension_cleant[-train_index, ]

#train for svm and knn scaled
train_scaled <- hypertension_scaled[train_index, ]
test_scaled <- hypertension_scaled[-train_index, ]


dim(train_data)
dim(test_data)

cat("\n========================\n")
dim(train_scaled)
dim(test_scaled)
```



#3.3.3. Report the class distribution (counts and percentages) in both training and testing sets.
```{r}
# Training Set
cat("Training Set:\n")
train_tab <- table(train_data$CVD)
print(train_tab)
cat("Percentages (%):\n")
print(round(prop.table(train_tab) * 100, 0))  # Changed from 2 to 0


cat("\n========================\n")

# Testing Set
cat("Testing Set:\n")
test_tab <- table(test_data$CVD)
print(test_tab)
cat("Percentages (%):\n")
print(round(prop.table(test_tab) * 100, 0))  # Changed from 2 to 0
```

3.4 Task 4: TrainControl Scheme (10 Marks)

#3.4.1. Define a trainControl scheme with the following specifications:
```{r}
ctrl <- trainControl(
method = "cv",              # 10-fold cross-validation
number = 10,                # 10 folds
savePredictions = "final",  # Save predictions
classProbs = TRUE,          # Enable probability predictions
verboseIter = FALSE,        # ROC, Sens, Spec

summaryFunction = twoClassSummary, # Suppress iteration output
)
```


#3.4.2. Explain why cross-validation is important for model evaluation.
Reduces overfitting by evaluating model on multiple unseen folds
Provides more stable and reliable performance estimates than a single train-test split
Allows hyperparameter tuning without leaking test data
Maximizes use of limited training data


#3.4.3. Save this control object to use across all models.
For fair comparison


3.5 Task 5: Model Training- Logistic Regression (15 Marks)

#3.5.1. Train a Logistic Regression model using caret::train() with method "glm".
```{r}
log_model <- train(
  
CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + physical_activity + 
      family_history + stress_level + sleep_hours,

data = train_data,
method = "glm",
family = binomial,
trControl = ctrl
)
```



#3.5.3. Extract and report the model coefficients.
```{r}
summary(log_model$finalModel)$coefficients
```


#3.5.4a) Calculate Odds Ratios (OR) for at least 3 significant predictors: OR = eβ
```{r}
# Calculate coefficients and odds ratios
coefficients <- coef(log_model$finalModel)
odds_ratios <- exp(coefficients)

# Create dataframe and sort
odds_df <- data.frame(
  Variable = names(odds_ratios),
  Odds_Ratio = round(odds_ratios, 3)
)

# Sort by Odds Ratio descending (largest to smallest)
odds_df <- odds_df[order(-odds_df$Odds_Ratio), ]

print(odds_df)
```


#3.5.4b)odds table
```{r}
odds_table <- data.frame(
  Variable = names(odds_ratios),
  Coefficient = round(coefficients, 4),
  Odds_Ratio = round(odds_ratios, 3)
)

odds_table <- odds_table[order(-abs(odds_table$Coefficient)), ]

odds_table
```


#3.5.4b)Bar plot of Odds Ratios
```{r}
# Remove intercept for better visualization
odds_table <- odds_table[odds_table$Variable != "(Intercept)", ]

# Sort by Coefficient absolute value
odds_table <- odds_table[order(-abs(odds_table$Coefficient)), ]


ggplot(odds_table, aes(x = reorder(Variable, Odds_Ratio), y = Odds_Ratio, 
                       fill = Odds_Ratio > 1)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1)+
  labs(title = "Odds Ratios for CVD Predictors",
       subtitle = "Values > 1 indicate risk factors, < 1 indicate protective factors",
       x = "Predictor Variable",
       y = "Odds Ratio") +
  scale_fill_manual(values = c("TRUE" = "coral", "FALSE" = "steelblue"),
                    labels = c("Risk Factor", "Protective Factor")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.y = element_text(size = 10)) +
  geom_text(aes(label = round(Odds_Ratio, 2)), hjust = -0.2, size = 3)
```

#3.5.5. Interpret the Odds Ratios in the context of CVD risk.
SLIDE


#3.5.2a) Alternatively, fit a baseline model using base R glm() function.
```{r}
log_modelb <- glm(
  CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + physical_activity + 
      family_history + stress_level + sleep_hours,
  
  data = train_data,
  family = binomial
)

summary(log_modelb)
```

#3.5.2b) Calculate Odds Ratios (OR) for at least 3 significant predictors: OR = eβ
```{r}
# Extract coefficients
coef_vals <- coef(log_modelb)

# Convert to odds ratios
odds_ratios <- exp(coef_vals)

# Get p-values from model summary
p_values <- summary(log_modelb)$coefficients[, 4]

# Create clean table
odds_table <- data.frame(
  Variable = names(coef_vals),
  Coefficient = round(coef_vals, 4),
  Odds_Ratio = round(odds_ratios, 3),
  P_value = round(p_values, 4)
)

# Remove row names
rownames(odds_table) <- NULL

# Sort by importance 
odds_table <- odds_table[order(-abs(odds_table$Coefficient)), ]

odds_table
```



3.6 Task 6: Model Training- KNN (15 Marks)

#3.6.1. Train a K-Nearest Neighbors model using caret::train() with method "knn".
#3.6.2a. Tune the parameter k (number of neighbors) using the grid: k = 1, 3, 5, ...,21.
```{r}
knn_grid <- data.frame(k = seq(1, 21, by = 2))

ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

knn_model <- train(
  CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + physical_activity + 
      family_history + stress_level + sleep_hours,
  
  data = train_scaled,
  method = "knn",
  tuneGrid = knn_grid,
  trControl = ctrl,
  metric = "ROC"
)

knn_model$results
```



#3.6.2a.plot KNN model
```{r}
plot(knn_model)
```


#3.6.3. Important: Use the scaled dataset for KNN (distance-based algorithm).




#3.6.4. Report the optimal k value selected by cross-validation.
```{r}
knn_model$bestTune
```


#3.6.5. Plot accuracy vs. k to visualize the tuning results.
```{r}
ggplot(knn_model$results, aes(x = k, y = ROC)) +
  geom_line() +
  geom_point(size = 2) +
  labs(title = "KNN Model Tuning (ROC vs k)",
       x = "Number of Neighbors (k)",
       y = "ROC AUC") +
  theme_minimal()
```




3.7 Task 7: Model Training- SVM (15 Marks)

#3.7.1. Train a Support Vector Machine model using caret::train() with method "svmRadial".
#3.7.2. Tune the hyperparameters:C (cost): 0.25, 0.5, 1, 2, 4, sigma: 0.01, 0.05, 0.1, 0.5
#3.7.3. Important: Use the scaled dataset for SVM.
```{r}
svm_model <- train(
  CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + physical_activity + 
      family_history + stress_level + sleep_hours,

data = train_scaled, # Use scaled data!
method = "svmRadial",
trControl = ctrl,
tuneGrid = expand.grid(
C = c(0.25, 0.5, 1, 2, 4),
sigma = c(0.01, 0.05, 0.1, 0.5)
))
  
svm_model$results
```


#3.7.4a. Report the optimal C and sigma values.
```{r}
 svm_model$bestTune
```

#3.7.5. Explain the role of the kernel function in SVM.
A kernel function is a mathematical tool that allows SVM to learn non-linear decision boundaries without explicitly transforming the data into higher dimensions.



#3.7.4b.Plot Heatmap of C vs Sigma for SVM
```{r}
# Extract SVM tuning results
svm_results <- svm_model$results

# Create heatmap
ggplot(svm_results, aes(x = factor(C), y = factor(sigma), fill = ROC)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(ROC, 3)), size = 4, color = "black") +
  scale_fill_gradient(low = "lightyellow", high = "darkred", 
                      name = "ROC AUC",
                      limits = c(min(svm_results$ROC), max(svm_results$ROC))) +
  labs(title = "SVM Hyperparameter Tuning Heatmap",
       subtitle = "ROC AUC by Cost (C) and Sigma (γ)",
       x = "Cost Parameter (C)",
       y = "Sigma (γ) - Kernel Width") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        plot.subtitle = element_text(hjust = 0.5),
        axis.text = element_text(size = 11),
        legend.position = "right")

# Save the plot
ggsave("svm_heatmap.png", width = 8, height = 6, dpi = 300)
```




3.8 Task 8: Model Training- Random Forest (15 Marks)

#3.8.1. Train a Random Forest model using caret::train() with method "rf".
#3.8.2. Tune the parameter mtry (features per split): 2, 4, 6, 8.
#3.8.3. Set ntree = 500 (number of trees).
```{r}
 rf_model <- train(
   CVD ~ age + gender + bmi + blood_pressure + cholesterol + 
      glucose + smoking + alcohol + physical_activity + 
      family_history + stress_level + sleep_hours,
  
  data = train_data, # Original data (not scaled)
method = "rf",
trControl = ctrl,
tuneGrid = expand.grid(mtry = c(2, 4, 6, 8)),
ntree = 500
)
rf_model$results 
```


#3.8.4. Report the optimal mtry value.
```{r}
rf_model$bestTune
```


#3.8.5. Extract and plot variable importance.
```{r}
varImp(rf_model)
plot(varImp(rf_model), main = "Random Forest: Variable
Importance")
```


#3.8.6a. Identify the top 5 most important predictors for CVD.
```{r}
rf_imp <- varImp(rf_model)

rf_imp_df <- rf_imp$importance
rf_imp_df$Variable <- rownames(rf_imp_df)
```


#3.8.6b.top 5 predictors
```{r}
top5 <- rf_imp_df[order(-rf_imp_df$Overall), ][1:5, ]
top5
```


#3.8.6b.plot Top 5 Important Predictors for CVD
```{r}
ggplot(top5, aes(x = reorder(Variable, Overall), y = Overall)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 5 Important Predictors for CVD",
       x = "Predictor",
       y = "Importance Score") +
  theme_minimal()
```


3.9 Task 9: Model Evaluation and Comparison (20 Marks)

#3.9.1. Predict the class and probabilities on the test set for all 4 models.
```{r}
log_pred <- predict(log_model, test_data)
log_probs <- predict(log_model, test_data, type = "prob")

knn_pred <- predict(knn_model, test_scaled)
knn_probs <- predict(knn_model, test_scaled, type = "prob")

svm_pred <- predict(svm_model, test_scaled)
svm_probs <- predict(svm_model, test_scaled, type = "prob")

rf_pred <- predict(rf_model, test_data)
rf_probs <- predict(rf_model, test_data, type = "prob")
 
```


#3.9.2a. For each model, generate a Confusion Matrix and report:
•Accuracy,•Sensitivity (Recall),•Specificity,•Precision (PPV),•F1-Score,•Kappa
#3.9.2a: Confusion Matrix and report (logistic regression )
```{r}
confusionMatrix(log_pred, test_data$CVD, positive = "Yes")
```

#3.9.2b: Confusion Matrix and report (KNN )
```{r}
confusionMatrix(knn_pred, test_scaled$CVD, positive = "Yes")
```


#3.9.2C: Confusion Matrix and report (SVM )
```{r}
confusionMatrix(svm_pred, test_scaled$CVD, positive = "Yes")
```


#3.9.2d: Confusion Matrix and report (RF )
```{r}
confusionMatrix(rf_pred, test_data$CVD, positive = "Yes")
```


#3.9.3. Plot ROC curves for all 4 models on the same graph.
#3.9.3a) get roc
```{r}
log_roc <- roc(test_data$CVD, log_probs$Yes)
knn_roc <- roc(test_data$CVD, knn_probs$Yes)
svm_roc <- roc(test_data$CVD, svm_probs$Yes)
rf_roc <- roc(test_data$CVD, rf_probs$Yes)
```


#3.9.3b)Plot ROC curves for all 4 models on the same graph.
```{r}
plot(log_roc, col = "blue", lwd = 2, main = "ROC Comparison")
 lines(knn_roc, col = "green", lwd = 2)
  lines(svm_roc, col = "red", lwd = 2)
  lines(rf_roc, col = "purple", lwd = 2)
  legend("bottomright",
         legend = c("Logistic", "KNN", "SVM", "RF"),
col = c("blue", "green", "red", "purple"),
lwd = 2)
```



#3.9.4. Calculate and compare AUC values for all models.
```{r}
auc_values <- data.frame(
  Model = c("Logistic", "KNN", "SVM", "Random Forest"),
AUC = c(as.numeric(auc(log_roc)),
as.numeric(auc(knn_roc)),
as.numeric(auc(svm_roc)),
as.numeric(auc(rf_roc)))
)
print(auc_values)
```


#3.9.5a) Create a comparison table summarizing all metrics.
```{r}
model_performance <- data.frame(
  Model = c("Logistic Regression", "KNN", "SVM", "Random Forest"),
  auc_values
)
model_performance <- model_performance[order(-model_performance$AUC), ]
model_performance
```





#3.9.6. Rank the models from best to worst based on AUC
```{r}
ggplot(model_performance, aes(x = reorder(Model, AUC), y = AUC)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Model Comparison Based on AUC",
       x = "Model",
       y = "AUC Score") +
  theme_minimal()
```


#3.9.5B) Create a comparison table summarizing all metrics.
```{r}
# Plot all ROC curves
plot(log_roc, col = "blue", lwd = 2, main = "ROC Comparison")
lines(knn_roc, col = "green", lwd = 2)
lines(svm_roc, col = "red", lwd = 2)
lines(rf_roc, col = "purple", lwd = 2)
legend("bottomright",
       legend = c("Logistic", "KNN", "SVM", "RF"),
       col = c("blue", "green", "red", "purple"),
       lwd = 2)

# AUC Comparison
auc_values <- data.frame(
  Model = c("Logistic", "KNN", "SVM", "Random Forest"),
  AUC = c(as.numeric(auc(log_roc)),
          as.numeric(auc(knn_roc)),
          as.numeric(auc(svm_roc)),
          as.numeric(auc(rf_roc)))
)
```

#3.9.2. For each model, generate a Confusion Matrix and report:
```{r}
 
# ============================================
# Function to extract all metrics from confusion matrix
# ============================================
extract_all_metrics <- function(cm, probs, actual, model_name) {
  # Calculate AUC
  roc_obj <- roc(actual, probs$Yes)
  auc_value <- as.numeric(auc(roc_obj))
  
  # Extract metrics from confusion matrix
  metrics <- data.frame(
    Model = model_name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    Precision = round(cm$byClass["Precision"], 4),
    F1_Score = round(cm$byClass["F1"], 4),
    Kappa = round(cm$overall["Kappa"], 4),
    AUC = round(auc_value, 4)
  )
  
  return(metrics)
}

# ============================================
# Generate confusion matrices and extract metrics
# ============================================

# Logistic Regression
cm_log <- confusionMatrix(log_pred, test_data$CVD, positive = "Yes")
metrics_log <- extract_all_metrics(cm_log, log_probs, test_data$CVD, "Logistic Regression")

# KNN
cm_knn <- confusionMatrix(knn_pred, test_scaled$CVD, positive = "Yes")
metrics_knn <- extract_all_metrics(cm_knn, knn_probs, test_scaled$CVD, "KNN")

# SVM
cm_svm <- confusionMatrix(svm_pred, test_scaled$CVD, positive = "Yes")
metrics_svm <- extract_all_metrics(cm_svm, svm_probs, test_scaled$CVD, "SVM")

# Random Forest
cm_rf <- confusionMatrix(rf_pred, test_data$CVD, positive = "Yes")
metrics_rf <- extract_all_metrics(cm_rf, rf_probs, test_data$CVD, "Random Forest")

# ============================================
# Combine all metrics into one table
# ============================================
comparison_table <- rbind(metrics_log, metrics_knn, metrics_svm, metrics_rf)

# ============================================
# Display the table
# ============================================
print(comparison_table)

# ============================================
# Save to CSV
# ============================================
write.csv(comparison_table, "model_comparison_metrics.csv", row.names = FALSE)

# ============================================
# Create a formatted table for presentation
# ============================================
library(knitr)
kable(comparison_table, 
      caption = "Model Performance Comparison - All Metrics",
      col.names = c("Model", "Accuracy", "Sensitivity", "Specificity", 
                    "Precision", "F1-Score", "Kappa", "AUC"),
      digits = 4)
```



3.10 Task 10: Interpretation and Recommendations (15 Marks)

#3.10.1. Based on your results, answer the following:
#3.10.1a) Which model performed best? Justify using metrics.
Logistic regression.
Highest AUC (0.85)
Highest sensitivity (28.99%)
Highest kappa (0.351)
Best balanced accuracy (63.47%)
Interpretable odds ratios for clinical insights



#3.10.1b)Which model performed worst? Explain possible reasons.
Random forest
2nd lowest AUC (0.790)
sensitivity (0%)
kappa (0.351)
Best balanced accuracy (63.47%)
Interpretable odds ratios for clinical insights

#3.10.1c)Was feature scaling necessary for all models? Explain.
feature scaling was nescessary for KNN,SVM Which are distance based model but not necessary for random forest a treebased model


#3.10.1d)Did cross-validation improve model reliability? Explain.
Yes cross validation avoided overfitting, and uniformity among all the models ensuarig reliabilty.


#3.10.2. For the best model, identify the top 3 risk factors for CVD. 
```{r}
# Remove intercept
odds_table <- odds_table[odds_table$Variable != "(Intercept)", ]

# Sort by absolute coefficient (model importance)
odds_table <- odds_table[order(-abs(odds_table$Coefficient)), ]

# Take top 5
top5_coef <- head(odds_table, 5)

# Plot
ggplot(top5_coef, aes(x = reorder(Variable, Odds_Ratio), y = Odds_Ratio, 
                       fill = Odds_Ratio > 1)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", linewidth = 1) +
  labs(title = "Top 5 Most Important CVD Predictors",
       subtitle = "Based on Logistic Regression Coefficient Magnitude",
       x = "Predictor Variable",
       y = "Odds Ratio") +
  scale_fill_manual(values = c("TRUE" = "coral", "FALSE" = "steelblue"),
                    labels = c("Risk Factor", "Protective Factor")) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        axis.text.y = element_text(size = 11)) +
  geom_text(aes(label = round(Odds_Ratio, 2)), hjust = -0.2, size = 4)
```



#3.10.3. Discuss the trade-offs between model accuracy and interpretability.
Logistic regression: lower complexity, HIGH interpretability, GOOD accuracy
• Random forest: higher accuracy, LOW interpretability (no direction)
• SVM: moderate accuracy, VERY LOW interpretability (black box)



#3.10.4. Provide recommendations for healthcare practitioners based on your findings.
(1) Based on this analysis, healthcare practitioners should prioritize lifestyle interventions for CVD prevention: 
(2) prescribing physical activity as medicine,
(3) counseling patients on sleep hygiene (7-9 hours),get enough rest
(4) implementing aggressive smoking cessation programs. 
(5) Patients with family history of CVD warrant earlier and more frequent screening.


#3.10.5. Discuss any limitations of your analysis (e.g., observational data, potential confounders).
Observational data,Class imbalance ,Data limitations,Model limitations,Clinical caution,Outliers








