---
title: "ML ASS 4 UL PENGUINS"
author: "KIBIDI"
date: "2026-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Course: PROFFESIONAL CERTIFICATE IN DATA ANALYSIS,  MACHINE LEARNING AND AI
Instructor: Lumumba W. Victor
Email: lumumbavictor172@gmail.com

student: KIBIDI LERENCER
Reg no: CDAM/0047/2026
Email: kabeyi24leren@gmail.com

Date Issued: April 13, 2026


#set working directory
```{r}
setwd("C:/Users/h6p/Desktop/submit")
```


3.1 Task 1: Data Import and Exploration (10 Marks)
#3.1.1. Load all necessary libraries (tidyverse, palmerpenguins, cluster, factoextra).
```{r}
library(tidyverse)      # Data manipulation andvisualization
library(palmerpenguins) # Built-in penguins dataset
library(cluster)        # Silhouette analysis
library(factoextra)     # PCA and clustering visualization
library(ggplot2)        # Advanced plotting
library(dplyr)          # Data wrangling
library(NbClust)
library(mclust)
```

#3.1.2. Load th epenguins dataset using data("penguins", package = "palmerpenguins").
```{r}
data("penguins", package = "palmerpenguins")
```


#3.1.3a) Display the structure    
```{r}
 str(penguins)
```


#3.1.3b)summary statistics
```{r}
summary(penguins)
```

#3.1.4a) Check for missing values and report the count and percentage per variable.
```{r}
colSums(is.na(penguins))
sum(is.na(penguins)) / (nrow(penguins) * ncol(penguins)) * 100
```

#3.1.4)summary of missing data
```{r}
missing_summary <- data.frame(
  Variable = names(penguins),
  Missing_Count = colSums(is.na(penguins)),
  Missing_Percent = colMeans(is.na(penguins)) * 100
)

missing_summary
```

#3.1.5. Create a clean dataset by removing rows with missing values: .
```{r}
penguins_clean <-na.omit(penguins)
dim(penguins_clean)
```
#3.1.5b)verify penguin clean has no  missing data
```{r}
missing_summary <- data.frame(
  Variable = names(penguins_clean),
  Missing_Count = colSums(is.na(penguins_clean)),
  Missing_Percent = colMeans(is.na(penguins_clean)) * 100
)

missing_summary
```



3.2 Task 2: Data Preprocessing (15 Marks)

#3.2.1. Select only numeric morphological variables for analysis:
• bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g
```{r}
numeric_vars <- c("bill_length_mm", "bill_depth_mm","flipper_length_mm", "body_mass_g")
 penguins_numeric <- penguins_clean[, numeric_vars]
```


#3.2.2. Scale these variables to mean = 0 and standard deviation = 1 using scale().
#3.2.3. Create a new dataframe penguins_scaled containing only the scaled numeric variables.
```{r}
 penguins_scaled <- as.data.frame(scale(penguins_numeric))
```



#3.2.4. Verify scaling 
```{r}
colMeans(penguins_scaled) # Should be ~0
apply(penguins_scaled, 2, sd) # Should be ~1
```

#3.2.5. Document why scaling is critical for K-Means and PCA
K-Means and PCA use Euclidean distance; unscaled variables with larger ranges (e.g., body_mass_g) would dominate the
analysis



3.3 Task 3: Exploratory Data Visualization (10 Marks)

#3.3.1. Create a scatterplot matrix of the four scaled numeric variables, coloredby the original species variable 
```{r}
 pairs(penguins_numeric,
       col = as.numeric(penguins_clean$species),
pch = 19, cex = 0.7,
main = "Penguin Morphology: Pairs Plot")
```

#3.3.2. Generate boxplots for each numeric variable grouped by species to visualize between species differences.
```{r}
penguins_clean %>%
  pivot_longer(cols = all_of(numeric_vars),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = species, y = value, fill = species)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Morphological Variables by Species")
```

#3.3.3. Compute and report the correlation matrix of the four numeric variables.
```{r}
 cor(penguins_numeric)
```





#3.3.4. Discuss which variables appear most useful for distinguishing groups



3.4 Task 4: K-Means: Finding Optimal K (10 Marks)

#3.4.1. Apply the elbow method: run K-Means for k = 1 to 10 and plot total within-clustersum of squares (WSS) vs. k.
```{r}
 set.seed(2026)

# Elbow method
wss <- map_dbl(1:10, ~{
  kmeans(penguins_scaled, centers = .x, nstart = 25)$tot.withinss
})

plot(1:10, wss, type = "b", pch = 19,
      xlab = "Number of Clusters (k)",
      ylab = "Total Within-Cluster Sum of Squares",
      main = "Elbow Method for Optimal k")
```



#3.4.2. Apply silhouette analysis: compute average silhouette width for k = 2 to 10 andplot results.
```{r}
# Silhouette analysis
sil_width <- map_dbl(2:10, ~{
  km <- kmeans(penguins_scaled, centers = .x, nstart = 25)
  silhouette(km$cluster, dist(penguins_scaled)) %>%
    summary() %>%
    `[[`("avg.width")
})

plot(2:10, sil_width, type = "b", pch = 19,
      xlab = "Number of Clusters (k)",
      ylab = "Average Silhouette Width",
      main = "Silhouette Analysis for Optimal k")
```


#3.4.2b. 
```{r}
ggplot(penguins_scaled, aes(x = bill_length_mm, y = bill_depth_mm, 
                            color = penguins_clean$cluster)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "K-Means Clusters: First Two Scaled Variables",
       x = "Bill Length (scaled)", y = "Bill Depth (scaled)") +
  theme_minimal()
```

#3.4.3. Based on both methods, justify your choice of optimal k.
both elbow mwetod amd silhouette analysis identify a achangea at 3 thus the optimal k is 3


#3.4.4. Report the total WSS and average silhouette width for your chosen k
```{r}
# Set seed for reproducibility
set.seed(2026)

# Fit K-Means with k=3
k_opt <- 3
km_model <- kmeans(penguins_scaled, centers = k_opt, nstart = 25)

# Extract Total WSS
total_wss <- km_model$tot.withinss
cat("Total Within-Cluster Sum of Squares (WSS):", round(total_wss, 2), "\n")

# Extract Average Silhouette Width
sil <- silhouette(km_model$cluster, dist(penguins_scaled))
avg_sil_width <- mean(sil[, 3])
cat("Average Silhouette Width:", round(avg_sil_width, 3), "\n")
```


3.5 Task 5: K-Means: Model Fitting and Cluster Profiling (15 Marks)
#3.5.1. Fit the final K-Means model using your optimal k and nstart = 25 for stability.
```{r}
set.seed(2026)
k_opt <- 3
km_model <- kmeans(penguins_scaled, centers = k_opt, nstart =25)
```


#3.5.2. Add the cluster assignments to the original penguins_clean dataframe.
```{r}
penguins_clean$cluster <- as.factor(km_model$cluster)
str(penguins_clean)
```


#3.5.3. Compute cluster centroids (mean of each scaled variable per cluster).
```{r}
 km_model$centers
```


#3.5.4. Profile each cluster by computing mean values of original (unscaled) variables per cluster.
#3.5.5. Create a table summarizing cluster characteristics (size, centroid values, dominant species if known).
```{r}
 penguins_clean %>%
  group_by(cluster) %>%
  summarise(across(all_of(numeric_vars), mean, na.rm = TRUE),
            n = n())
```





#3.5.6. Visualize clusters using a scatter plot of the first two scaled variables, colored by cluster assignment.
```{r}
fviz_cluster(km_model, data = penguins_scaled,
             geom = "point", ellipse.type = "norm",
             palette = "jco", ggtheme = theme_minimal(),
             main = "K-Means Clustering of Penguin Morphology")
```

#COUNTS
```{r}
table(penguins_clean$cluster)
```


3.6 Task 6: Principal Component Analysis (PCA) (15 Marks)

#3.6.1. Perform PCA on the scaled numeric variables using prcomp().
```{r}
pca_result <- prcomp(penguins_scaled, center = TRUE, scale. =TRUE)

# Summary and variance
summary(pca_result)
pca_result$sdev^2 / sum(pca_result$sdev^2) # Proportion of variance
```

#3.6.2. Generate and interpret the scree plot: report the proportion of variance explained by each PC.
```{r}
fviz_eig(pca_result, addlabels = TRUE, ylim = c(0, 80),
         main = "Scree Plot: Variance Explained by PCs")
```

#3.6.3. Determine how many principal components are needed to explain at least 90% of total variance.
```{r}
var_prop <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cum_var <- cumsum(var_prop)
pcs_90 <- which(cum_var >= 0.90)[1]
cat("PCs needed for ≥90% variance:", pcs_90, "\n")
cat("Cumulative variance explained:", round(cum_var[pcs_90]*100, 2), "%\n")
```





#3.6.4. Extract and interpret the loadings (rotation matrix) for PC1 and PC2.
```{r}
pca_result$rotation[, 1:2]
```




#3.6.5. Create a biplot showing observations and variable loadings in PC1-PC2 space
```{r}
fviz_pca_biplot(pca_result,
                label = "var", 
                habillage = "none",
                addEllipses = FALSE, 
                col.var = "black",
                repel = TRUE) +
  ggtitle("PCA Biplot: PC1 vs PC2")
```




3.7 Task 7: Combining K-Means and PCA (15 Marks)

#3.7.1. Project the scaled data onto the first two principal components.
```{r}
pca_scores <- as.data.frame(pca_result$x[, 1:2])
colnames(pca_scores) <- c("PC1", "PC2")
```



#3.7.2. Re-run K-Means clustering on the first two PCs only (using your optimal k).
```{r}
set.seed(2026)
km_pca <- kmeans(pca_scores, centers = k_opt, nstart = 25)

```



#3.7.3. Compare cluster assignments from PCA-reduced data vs. original 4-dimensionaldata (use a confusion matrix).
```{r}
 table(FullSpace = km_model$cluster, PCspace = km_pca$cluster)
```




#3.7.4. Create a scatter plot of PC1 vs. PC2, colored by:
• K-Means cluster assignment (from full data)
• Original species (for validation reference only)
```{r}
pca_scores$cluster_km <- as.factor(km_model$cluster)
  ggplot(pca_scores, aes(x = PC1, y = PC2, color = cluster_km))+
  geom_point(size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(title = "PC1 vs PC2: Colored by K-Means Clusters (Full Data)")
```

#3.7.5. Discuss whether PCA preserves the cluster structure found in the full-dimensional space.
```{r}
pca_scores$species <- penguins_clean$species
 ggplot(pca_scores, aes(x = PC1, y = PC2, color = species)) +
   geom_point(size = 2, alpha = 0.7) +
   theme_minimal() +
   labs(title = "PC1 vs PC2: Colored by Species (Reference)")
```






3.8 Task 8: Cluster Validation and Quality Assessment (10 Marks)

#3.8.1a) Compute internal validation metrics for your K-Means solution:
```{r}
sil <- silhouette(km_model$cluster, dist(penguins_scaled))
   mean(sil[, 3]) # Average silhouette width
   plot(sil, col = 1:k_opt, border = NA, main = "Silhouette Plot")
```


#3.8.1b)Average silhouette width•Dunn index (using cluster::dunn())
```{r}
#library(cluster)
 #dunn_index <- dunn(penguins_scaled, cluster = km_model$cluster)
 #dunn_index
```


#3.8.1b)Average silhouette width•Dunn index (using cluster::dunn()Manual Dunn Index function
```{r}
dunn_manual <- function(data, clusters) {
  # Get unique clusters
  cluster_ids <- unique(clusters)
  k <- length(cluster_ids)
  
  # Calculate intra-cluster distances (within cluster)
  intra_dists <- c()
  for(i in 1:k) {
    cluster_points <- data[clusters == cluster_ids[i], ]
    if(nrow(cluster_points) > 1) {
      dist_matrix <- as.matrix(dist(cluster_points))
      intra_dists <- c(intra_dists, max(dist_matrix[upper.tri(dist_matrix)]))
    } else {
      intra_dists <- c(intra_dists, 0)
    }
  }
  
  # Calculate inter-cluster distances (between clusters)
  inter_dists <- c()
  for(i in 1:(k-1)) {
    for(j in (i+1):k) {
      cluster_i <- data[clusters == cluster_ids[i], ]
      cluster_j <- data[clusters == cluster_ids[j], ]
      
      # Find minimum distance between clusters
      dist_matrix <- as.matrix(dist(rbind(cluster_i, cluster_j)))
      n_i <- nrow(cluster_i)
      inter_min <- min(dist_matrix[1:n_i, (n_i+1):ncol(dist_matrix)])
      inter_dists <- c(inter_dists, inter_min)
    }
  }
  
  # Dunn = min(inter) / max(intra)
  dunn_value <- min(inter_dists) / max(intra_dists)
  return(dunn_value)
}

# Calculate Dunn index
dunn_index <- dunn_manual(penguins_scaled, km_model$cluster)
print(dunn_index)
```


#3.8.1c)Calinski-Harabasz index (using factoextra::fviznbclust())
```{r}
 #fviz_nbclust(penguins_scaled, kmeans, method = "ch", k.max =10) +
#  labs(title = "Calinski-Harabasz Index vs k")
```


#3.8.1c)Calinski-Harabasz index (using factoextra::fviznbclust())
```{r}
# Calculate Calinski-Harabasz index for k=2 to 10
ch_results <- NbClust(data = penguins_scaled,
                      method = "kmeans",
                      index = "ch",
                      min.nc = 2,
                      max.nc = 10)

# View results
print(ch_results$All.index)
```


#3.8.2. Report these metrics and interpret their values (what indicates good clustering?)
Silhouette (0.446) indicates weak-to-reasonable structure.   
Dunn index (0.037) suggests moderate separation.
CH index (427.8) supports the 3-cluster solution despite a peak at k=2.

.
#3.8.3. Optional: Compare metrics for k = 2,3,4 to confirm your choice of optimal k.
k=3 ffrom elbow and Silhouette method


#3.8.4. Discuss limitations of internal validation metrics
Silhouette assumes spherical clusters - penguin data may have elongated shapes
Dunn index sensitive to outliers - can be misleading with noisy data
CH index favors convex clusters - may not capture complex natural groupings
All internal metrics lack ground truth - should be complemented with biological validation




Interpretation notes:Silhouette: >0.5 = reasonable, >0.7 = strong structure
#- Dunn: higher = better separated, more compact clusters
#- CH: higher = better clustering




3.9 Task 9: Interpretation and Biological Insights (15 Marks)


#3.9.2. Compare your discovered clusters to the known species labels (for validation only):
compute a contingency table and discuss agreement.
#3.9.1. Based on your clustering results, describe the morphological profile of each cluster ieCluster 1: small bill.
```{r}
table(penguins_clean$cluster, penguins_clean$species)

library(mclust)
adjustedRandIndex(penguins_clean$cluster, penguins_clean$
species)

```


#3.9.3. Interpret the PCA loadings: which original variables contribute most to PC1 andPC2?
```{r}
 
loadings_df <- as.data.frame(pca_result$rotation[, 1:2])
loadings_df$variable <- rownames(loadings_df)
 print(loadings_df)

```

What biological meaning might these components have?

#3.9.4. Discuss whether the clusters you found align with known biological groupings (species,island, sex).
they did 3 clusters corresponding to Adelie, Chinstrap, Gentooin population size, features

#3.9.5. Propose one ecological or conservation question that could be addressed using theseunsupervised learning results.
# Example interpretation:
If flipper_length and body_mass have high positive loadings on PC1,
# then PC1 represents "overall body size"




3.10 Task 10: Report Synthesis and Recommendations (15 Marks)

#3.10.1. Summarize your complete workflow: data preparation, K-Means, PCA, validation.
1.Ideentifying the main objective of the assignment
2.Data Preparation
3.K-Means Clustering:
4.Principal Component Analysis:
5. Validation
6.interpretation 



#3.10.2a. Answer: Did K-Means discover meaningful groups in the penguin data? 
yes


#3.10.2b. Justifyusing metrics and visualizations.
scatter plot

#3.10.3a. Answer: Was dimensionality reduction via PCA useful? 
 PCA preserved cluster structure almost perfectly while reducing 4 dimensions to 2, enabling effective visualization and faster computation.


#3.10.3b.Did it preserve clusterstructure?
yes

#3.10.4. Discuss trade-offs: When would you prefer clustering in original space vs. PCAreduced space?
reduced space faster to analyse

#3.10.5a. Provide recommendations for a wildlife biologist: 
Scale data, good to remove ouliers, use multiplemrthods to determine k,


#3.10.5b.How could these unsupervised methods help in field studies where species labels are unknown?it
it would help classify those with similar features into a common clusters thus enabling labeling


#3.10.6. Discuss limitations: What assumptions do K-Means and PCA make? How might violations affect your results?
Silhouette assumes spherical clusters - penguin data may have elongated shapes
Dunn index sensitive to outliers - can be misleading with noisy data
CH index favors convex clusters - may not capture complex natural groupings
All internal metrics lack ground truth - should be complemented with biological validation



